---
title: "Data science session 3"
author: "Ina"
date: today
date-format: long
format: 
  html:
    self-contained: true

---

```{r}

library(tidyverse)
library(palmerpenguins)
library(gapminder)
```

## Discussion of week 2 exercises


## R Basics

Create new objects with the assignment operator `<-` 
```{r}

```

What you can do with an object depends on its type, e.g. basic arithmetic with numeric objects only

```{r}


```

To see the result of a computation that creates a new object, either type the name of the object to inspect it or enclose the operation in parentheses.

```{r}

```

Example of a more complex object name

```{r}


```

Example of a function: `seq()`

```{r}

```

Be careful with brackets and quotation marks

```{r}

```

## A little more on data manipulation

Create new objects from the penguins data

```{r}

```

Additional ways to summarise

```{r}


```

Other summary functions: `median()`, `min()`, `max()`

```{r}

```

To look at groups use `group_by()`

```{r}


```

To look at the number of observations in a group use `count()`

```{r}

```

## Plotting 

### Scatterplots for two continuous variables

Scatterplot example:

```{r}


```

### Add title and axis labels

Adding a meaningful title and axis labels makes your plot more informative and readable.

```{r}


```

## Plotting a categorical predictor of a continuous variable

What about species as a predictor of body mass?

```{r}


```

`geom_jitter()` produces a better result:

```{r}


```

## Bar / column graphs to plot quantities

Alternative approach: plot mean body mass by species using a bar / column graph with `geom_col()`:

```{r}


```

It is crucial to summarise first in order to only have one value per species, otherwise the plot won't show what you think:

```{r}


```


Note that there is also a `geom_bar()`. Have a look at the documentation: what is the difference between `geom_col()` and `geom_bar()`?

## Line graphs to plot changes over time

Line graph to plot the number of sightings per year
We first need to count the number of sightings

```{r}


```

Adding groups, e.g. to look at the number of sightings per year for each species:

```{r}



```

## Histograms to examine distributions

Use a histogram to look at a distribution (e.g. of bill depth).

```{r}


```

Pair a histogram with facets to look at groups:

```{r}


```

## Boxplots to examine distributions

- Line = median
- Lower edge of box = 1st quartile
- Upper edge of box = 3rd quartile
- Whiskers extend to largest / smallest value, maximally 1.5 interquartile ranges from edge of box
- Data points beyond the whiskers are deemed outliers and plotted individually

```{r}

```

Add colour fill:

```{r}


```

## Gapminder data

- How has life expectancy in Australia changed over time?
- What was the relationship between life expectancy and how wealthy a country is (gdpPercap) in 1952? In 2007?
- How did the mean life expectancy per continent change over time?
- Which continent had the highest variability in population across countries? Pick two years of your choice to examine.
